---
title: "Assignment 2"
author: "Jasper Ginn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{assignment2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

The code for this assignment is stored in the 'meerkat' package. It contains helper functions that are not exported and for which no visible documentation was written. Please see the code that I sent you for a detailed overview for each function.

## Part I

The code for the t-test can be found [here](https://github.com/JasperHG90/meerkat/blob/master/R/a2_ttest.R)

#### 1.

See the code I sent to you.

#### 2. 

```{r}
library(meerkat)
# Create two variables
CSFI <- c(2,5,5,6,6,7,8,9)
TFI <- c(1,1,2,3,3,4,5,7,7,8)

# Calculate t-test
set.seed(500)
tt1 <- meerkat::t_test(CSFI, TFI, variance_equal = TRUE, 
                       bootstrap_ssize = 0.7, 
                       R=1000)
tt1
```

The results indicate that there is no statistically different performance between pilots who use traditional flight instructions and pilots who are trained using computer-simulated flight instruction.

When we compare this to the built-in R function, we get:

```{r}
t.test(CSFI, TFI, var.equal = TRUE)
```

The t-test We observe that the 95\% confidence interval contains $0$. This means that we may conclude there is no effect.

#### 3.

See the code I sent to you

#### 4.

The print method of the `t_test()` function prints the information returned by this function in a neatly organized way. 

The `t_test()` function returns an S3 object containing:

- **inputs**: the x, y variables entered by the user. Also contains a boolean whether or not the test should be computed for equal variances.
- **summary_statistics**: summary statistics for x, y variables & pooled variance if using equal variances.
- **test**: t-statistic, degrees of freedom and p-value

```{r}
t1 <- t_test(CSFI, TFI)
names(t1)
```

## PART II

The code for the linear model can be found [here](https://github.com/JasperHG90/meerkat/blob/master/R/a2_linearmodel.R)

```{r}
data("gala")
```

#### 1.

```{r}
mod1 <- lm("Species ~ Area + Elevation + Endemics", data = gala)
summary(mod1)
```

```{r}
residuals_mod1 <- resid(mod1)
predicted_mod1 <- predict(mod1)

# Plot
plot(predicted_mod1, residuals_mod1)
```

#### 2.

```{r}
# Make subset of the variables
y <- as.matrix(gala[, "Species"])
# Create a formula
form <- formula("Species ~ Area + Elevation + Endemics")
# Create a model matrix
X <- model.matrix(form, gala)
# Apply the formula for the linear model
linmod <- solve( t(X) %*% X ) %*% t(X) %*% y
# Create the predicted values
yhat <- X %*% linmod
# Residuals
resid <- (y - yhat)
# Plot
plot(yhat, resid)
```

The results obtained are the same as those obtained from the `lm()` function

#### 3.

```{r}
library(meerkat)

gala_mod <- linear_model(form, gala)
gala_mod
```

```{r}
summary(gala_mod)
```

```{r}
plot(gala_mod)
```
